Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-15 article.Rmd
More citations

R and Python are two of the most commonly used languages for data analysis \[Piatesky 2014, Muenchen 2014-]\.^[Surveys and analysis of useage point out the growth of R and the continuing popularity of Python among "data science" generally. See, for example, ] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which includes both the predictive algorithm and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package containing various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.^[The gender package depends especially on R Core Team, *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria (2014). <http://www.R-project.org/>;  Hadley Wickham and Romain Francois, *dplyr: dplyr: a grammar of data manipulation*. R package version 0.2 (2014): <http://CRAN.R-project.org/package=dplyr>. The sources of the data are cited below.]

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by government entities, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years. since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as providing a significant correction to predictions based on Social Security Administration data.

@@

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

In conclusion, our algorith for predicting gender from first names combined with historical datasets in the Gender package provides a powerful new method for historians interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance historians always have to deal with the problem of scarcity in the sources. Our methods uses the abundant datasets to produce more information from scarce datasets. The temporal dimension of our work matches a historical question to a historical method, providing much improved over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make the predictions. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.

@@

@@

@@

>>>>

R and Python are two of the most commonly used languages for data analysis \[Piatesky 2014, Muenchen 2014-]\.^[Surveys and analysis of useage point out the growth of R and the continuing popularity of Python among "data science" generally.] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[**Python library names**] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages used in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which includes both the predictive algorithm and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package containing various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.^[The gender package depends especially on \[R Core Team 2014-, Wickham and Francois 2014]\ ]

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by government entities, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years, since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as providing a significant correction to predictions based on Social Security Administration data.

@@

The Social Security Administration (SSA) "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910 \[Social Security Administration 2014-b]\.] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[Shane Landrum has studie dthe history of birth registration in the United States \[Landrum 2014]\.  It is well-known among American historians that the Social Security Act was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

In conclusion, our algorithm for predicting gender from first names combined with historical datasets in the Gender package provides a powerful new method for historians interested in the study of gender. The digital turn has made available very large datasets, yet in the midst of abundance historians always have to deal with the problem of scarcity in the sources. Our methods uses the abundant datasets to produce more information from scarce datasets. The temporal dimension of our work matches a historical question to a historical method, providing much improved over simple, anachronistic lookup methods. This method is extensible to any time or place for which there is a suitable dataset of names from which to make the predictions. As we have demonstrated in our study of gatekeeping in the American historical profession, our method can reveal important details about gender even from datasets which do not otherwise include that information, and it can do so over long historical periods.

@@

Landrum, in progress. Landrum, Shane. "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950." PhD dissertation, Brandeis University, in progress.

@@

R Core Team 2014-. R Core Team, *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria, 2014. <http://www.R-project.org/>

Social Security Administration 2014-a. Social Security Administration, "Baby Names from Social Security Card Applications-National Level Data". Data.gov dataset ID US-GOV-SSA-338. Accessed 31 August 2014, http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data

Social Security Administration 2014-b. Social Security Administration, "Baby Names from Social Security Card Applications-Data by State and District of Columbia." Data.gov dataset ID US-GOV-SSA-339. Accessed 31 August 2014, http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-

@@

Wickham and Francois 2014-. Wickham, Hadley and Romain Francois, *dplyr: dplyr: a grammar of data manipulation*. R package version 0.2 (2014): <http://CRAN.R-project.org/package=dplyr>.
